---
title: "EcoTopic2_eBird"
author: "Stellah and Claire" 
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is me changing something... 
Stellah Change 

## R Markdown

Research Question: 
How is noise pollution related to population distribution of birds across NY state? 

Hypotheses: 

Increased proximity to noise pollution will be associated with a decreased bird population as noise pollution can disrupt breeding and communication of birds as well as increase stress levels. 

Increased volume of noise pollution will be associated with a decreased bird population as higher volumes of noise pollution are more disruptful to breeding and communication of birds as well as increasing stress. 


```{r}
#general definitions:
##NP = noise pollution 
```

Library
```{r}
#library any new packages we need to analyze 
#comment what the package will be used for 
library(dplyr)
library(terra)
library(sf)
library(maps)
```

Loading Data
```{r}
#load in data 
ny_raster_np_data <- rast('/Users/smarient/Documents/MacalesterCollege /Spring2026/BigDataEcology/EcoTopic2_NY/NY_rail_road_and_aviation_noise_2020.tif')

#load in ebird data 

bird_raw_data <- read.table('~/Library/CloudStorage/GoogleDrive-cwiley@macalester.edu/My Drive/1A: Spring 2026/Big Data in Ecology/EcoTopic2/Eco_Topic2_NY/ebd_US-NY_202001_202012_smp_relJan-2026/ebd_US-NY_202001_202012_smp_relJan-2026.txt', sep = "\t", header = TRUE, fill = TRUE)

head(bird_raw_data)

##taking out unneeded columns in ebird data 

bird_wrangled<- bird_raw_data%>%
  select(-LAST.EDITED.DATE, -TAXON.CONCEPT.ID, -EXOTIC.CODE, -BREEDING.CODE, -BREEDING.CATEGORY, -BEHAVIOR.CODE, -AGE.SEX, -COUNTRY, -COUNTRY.CODE, -STATE, -STATE.CODE, -OBSERVER.ID, -OBSERVER.ORCID.ID, -SAMPLING.EVENT.IDENTIFIER, -PROTOCOL.NAME, -PROTOCOL.CODE, -PROJECT.NAMES, -PROJECT.IDENTIFIERS, -APPROVED, -REVIEWED, -REASON, -HAS.MEDIA, -GROUP.IDENTIFIER, -NUMBER.OBSERVERS, -EFFORT.DISTANCE.KM, -EFFORT.AREA.HA, -DURATION.MINUTES)

head(bird_wrangled)

bird_nyc <- bird_wrangled%>%
  c("Bronx", "Kings", "New York", "Queens", "Richmond")

#preliminary map
plot(ny_raster_np_data,
  col = terrain.colors(100),
  main = "Noise Pollution Map",
  xlab = "Longitude",
  ylab = "Latitude",
  legend = TRUE)

#convert all zeroes to 45 decibels 
ny_raster_np_data[ny_raster_np_data == 0] <- 45
plot(ny_raster_np_data)
ny_raster_np_agg <- aggregate(ny_raster_np_data, fact = 5, fun = mean)
plot(ny_raster_np_agg, col = rev(terrain.colors(50)))
hist(ny_raster_np_data)
```

```{r}
nyc_npboundary <- ext(1800000, 1900000, 2100000, 2230000)
nyc_npcrop <- crop (ny_raster_np_agg, nyc_boundary)
plot(nyc_npcrop, col = rev(terrain.colors(50)))
```

```{r}
##NP -- DOT Transportation 2020, clipping NYS: Continental U.S. road, freight and passenger rail, and aviation noise 2020
#Link: https:/www.bts.gov/geospatial/national-transportation-noise-map 

##EBird -- NY State data for 2020
##Counties -- NY State country boundaries

##naming convention is all lowercase with underscores and use "data" at end for raw data 

#head() all of these datasets to get a feel for their contents 
```



NP Histogram
```{r}
#make a histogram of the distribution of decibels of NP in NY to understand the data

#are there clear bin lines that we can create to create an index for NP? Then could use that throughout our analysis to simplify the NP data -- if yes, explore with boxplots
```

NP aggregation by county & mapping
```{r}
#aggregate NP by county in NY 

#map decibels of NP across NY by county 
```

Bird aggregation by county investigation
```{r}
#basic map of all bird sightings in NY in 2020 to see if aggregating by county (which we suspect will be effective) is actually a good way to group the data

#preliminary map

initial_bird_map <- map_data("bird_nyc")

#aggregate bird sightings by county 

#make a histogram of the distribution of bird observations by county

#make a bivariate map of the bird count and NP by county 
```

Exploring overall linear relationship
```{r}
#create a scatterplot for the relationship between noise pollution and bird sightings (expecting a negative linear relationship)

#run a lm on this relationship 
```

Diving into specific species  
```{r}
#find the top 5 most common birds in the dataset and compare them 
##consider ecological research to determine choice 

#create boxplots for the relationship between noise pollution index and each species count (differentiate by color)
```

NYC Case Study 
```{r}
##make a histogram of the distribution of decibels of NP in NYC to understand the data

###may need to create a new index for this -- as NYC will be louder than the rest of NY

#map all bird sightings to familiarize with distribution across the city  
###maybe by county with all points on top of that -- group by (county) -- and fill with color based on count  

#create scatterplot for the relationship between NP and bird sightings 

##run a lm on this^^^ relationship

#create boxplots for the relationship between NP Index and bird count 


##dive into the 5 most common birds analyzed above (same species)

###create boxplots for the relationship between NP index and each species count (differentiate by color)
```

