---
title: "EcoTopic2_eBird"
author: "Stellah and Claire" 
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is me changing something... 
Stellah Change 

## R Markdown

Research Question: 
How is noise pollution related to population distribution of birds across NY state? 

Hypotheses: 

Increased proximity to noise pollution will be associated with a decreased bird population as noise pollution can disrupt breeding and communication of birds as well as increase stress levels. 

Increased volume of noise pollution will be associated with a decreased bird population as higher volumes of noise pollution are more disruptful to breeding and communication of birds as well as increasing stress. 


```{r}
#general definitions:
##NP = noise pollution 

#Planned Data to use

##NP -- DOT Transportation 2020, clipping NYS: Continental U.S. road, freight and passenger rail, and aviation noise 2020
#Link: https:/www.bts.gov/geospatial/national-transportation-noise-map 

##EBird -- NY State data for 2020
##Counties -- NY State country boundaries

##naming convention is all lowercase with underscores and use "data" at end for raw data 

#head() all of these datasets to get a feel for their contents 
```

Library
```{r}
#library any new packages we need to analyze 
#comment what the package will be used for 
library(dplyr) 
library(terra)
library(sf) #using spatial data
library(tmap) #mapping
library(ggplot2) #vizualizations 
```

Loading Data
```{r}
#load in data 
ny_raster_np_data <- rast('NY_rail_road_and_aviation_noise_2020.tif')

#load in ebird data 

bird_raw_data <- read.table('ebd_US-NY_202001_202012_smp_relJan-2026.txt', sep = "\t", header = TRUE, fill = TRUE)

##taking out unneeded columns in ebird data 

bird_wrangled<- bird_raw_data%>%
  select(-LAST.EDITED.DATE, -TAXON.CONCEPT.ID, -EXOTIC.CODE, -BREEDING.CODE, -BREEDING.CATEGORY, -BEHAVIOR.CODE, -AGE.SEX, -COUNTRY, -COUNTRY.CODE, -STATE, -STATE.CODE, -OBSERVER.ID, -OBSERVER.ORCID.ID, -SAMPLING.EVENT.IDENTIFIER, -PROTOCOL.NAME, -PROTOCOL.CODE, -PROJECT.NAMES, -PROJECT.IDENTIFIERS, -APPROVED, -REVIEWED, -REASON, -HAS.MEDIA, -GROUP.IDENTIFIER, -NUMBER.OBSERVERS, -EFFORT.DISTANCE.KM, -EFFORT.AREA.HA, -DURATION.MINUTES)

#changing lat and log to be numeric and filtering out nas 
bird_wrangled <- bird_wrangled %>%
  dplyr::mutate(
    LATITUDE = as.numeric(LATITUDE),
    LONGITUDE = as.numeric(LONGITUDE)
  )%>%
 dplyr::filter(!is.na(LATITUDE) & !is.na(LONGITUDE))

#filtering for just NYC
bird_nyc <- bird_wrangled%>%
  filter(COUNTY %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))
```

Starting with NP Data
```{r}
#preliminary map
plot(ny_raster_np_data,
  col = terrain.colors(100),
  main = "Noise Pollution Map",
  legend = TRUE)

hist(ny_raster_np_data)
#tons of zeros 

#convert all zeroes to 45 decibels -- which is the lower end of the np scale measured; these places will not have actually zero noise pollution, especially in NYC, so the zeros will be misrepresentation 

#ny_raster_np_data[ny_raster_np_data == 0] <- 45
plot(ny_raster_np_data)

ny_raster_np_agg <- aggregate(ny_raster_np_data, fact = 5, fun = mean)
plot(ny_raster_np_agg, col = rev(terrain.colors(50)))
hist(ny_raster_np_agg)

#making np boundary for NYC
nyc_npboundary <- ext(1800000, 1900000, 2100000, 2230000)
nyc_npcrop <- crop (ny_raster_np_agg, nyc_boundary)
plot(nyc_npcrop, col = rev(terrain.colors(50)))
```

NP Boxplots
```{r}
#are there clear bin lines that we can create to create an index for NP? Then could use that throughout our analysis to simplify the NP data -- if yes, explore with boxplots
```

Bird Data Investigation
```{r}
#making it a spatial file 
bird_spatial <- st_as_sf(bird_nyc, 
                         coords = c("LONGITUDE", "LATITUDE"), 
                         crs = 4326)

#preliminary map
bird_initial_map <- tm_shape(bird_spatial)+
  tm_dots("COUNTY")
bird_initial_map

#make a bar chart of the distribution of bird observations by county

ggplot(bird_spatial, aes(x=COUNTY)) + 
  geom_bar(fill="skyblue", color="black") +
  labs(title="Bird Sightings by County", x="County", y="Frequency") +
  theme_minimal()
```

Extract NP value for each bird point 
```{r}
bird_np_extraction <- extract(ny_raster_np_data, bird_spatial)


maybe bin them after scatterplot 

```


Exploring overall linear relationship
```{r}
#create a scatterplot for the relationship between noise pollution and bird sightings (expecting a negative linear relationship)

#run a lm on this relationship 
```

Diving into specific species  
```{r}
#find the top 5 most common birds in the dataset and compare them 
##consider ecological research to determine choice 

#create boxplots for the relationship between noise pollution index and each species count (differentiate by color)
```


